Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2020-09-03 journal paper/content/sec-intro.tex
A pass on the intro: compressed the model & algorithms, expanded on dynamic vs static strategies. Moved "discussion of the model" to the model section.

%We investigate these questions with
\xhdr{Our model: Competition game.} We consider a stylized duopoly model where two firms commit to exploration strategies and compete for a stream of consumers. We define a game in which two firms (\emph{principals}) simultaneously engage in exploration and compete for users (\emph{agents}). These two processes are interlinked, as exploration decisions are experienced by users and informed by their feedback. We need to specify several conceptual pieces: how the principals and agents interact, what is the machine learning problem faced by each principal, and what is the information structure. Each piece can get rather complicated in isolation, let alone jointly, so we strive for simplicity. Thus, the key features of our model are as follows:

%\begin{itemize}

\textbf{(i)} A new agent arrives in each round $t=1,2, \ldots$, and chooses among the two principals. The principal chooses an action (\eg a list of web search results to show to the agent), the user experiences this action, and reports a reward. All agents have the same ``decision rule" for choosing among the principals given the available information.

\textbf{(ii)} Each principal
%learning problem is as follows: 
faces a basic and well-studied version of the multi-armed bandit problem: 
for each arriving agent, it chooses from a fixed set of actions  (a.k.a. \emph{arms}) and receives a reward drawn independently from a fixed distribution specific to this action. The reward distributions are initially unknown.
% (but can be estimated over time from the data).

\textbf{(iii)} Principals simultaneously announce their bandit algorithms before round $1$, and cannot change them afterwards.  Each principal's objective is to maximize its market share (the fraction of users choosing this principal). Each principal only observes agents that chose this principal.
%, but may have access to each platforms' reputation score (more on this in    Section~\ref{sec:intro-discussion}).
%\end{itemize}

We draw on the rich literature on multi-armed bandits to compare bandit algorithms to one another in the absence of competition, so that we can talk about \emph{better} bandit algorithms in a principled way.%
\footnote{Comparisons between bandit algorithms are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. More on this in Preliminaries.} One baseline is the \emph{greedy algorithm} that does not explicitly explore; in the absence of competition, it is known to be terrible for a wide variety of problem instances.

We investigate several model variants within this framework, where we vary agents' decision rule and/or give a first-mover advantage to one of the principals. In all variants, agents have little or no information about other agents' choices and rewards.

@@

\xhdr{Theoretical results.}
%We endow agents with Bayesian rationality, a common modeling approach for a theoretical investigation.
We consider a basic Bayesian model. We posit that agents have a common Bayesian prior on reward distributions, and know the principals' algorithms. For simplicity, agents do not receive any information about the previous agents' choices and rewards. Each agent knows the round (s)he arrives in, computes the Bayesian-expected reward for each principal, and use these two numbers to decide which principal to choose. We refer to this variant as the \emph{\TheoryModel}.

@@

\ascomment{stopped here.}

@@

\ascomment{One downside of the story above: we may need to explain why we didn't simulate \SoftMaxRandom.}

\ascomment{The xhdr below is basically from Guy, with some rewordings by Alex.}

@@

\ascomment{"Econ interpretation" xhdrs fit incely in "our results", as opposed to a separate subsection. Because: we can explain the "inverted-U" stuff right after "theory", before we put the simulation results on the stack. And the "network effects" stuff goes nicely right after the numerical results, without a big subsection break.}

\subsection{Further discussion}
\label{sec:discussion}

@@

\ascomment{The next para makes a point that seems important (to me). Is this point clear?}

@@

\xhdr{Modeling.}
Our models are stylized in several important  respects. Firms compete only on the quality of service, rather than, say, pricing or the range of products. Agents are myopic: they do not worry about how their actions impact their future utility.\footnote{In particular, agents do not attempt to learn over time, to second-guess or game future agents, or to manipulate the principals' learning algorithms. This is arguably typical in practice, in part because one agent's influence tends to be relatively small.}  Various performance signals available to the users, from personal experience to word-of-mouth to consumer reports, are abstracted as persistent ``reputation scores", and further simplified to average performance over a time window. On the machine learning side, our setup is geared to distinguish between ``simple" vs. ``better" vs. ``smart" bandit algorithms; we are not interested in state-of-art algorithms for very realistic bandit settings.

We consider two extremes: a simple Bayesian model with full Bayesian rationality and no performance signals, and a simple frequentist model with reputation scores and no prior knowledge or beliefs. For the theoretical results, the ``no-performance-signals" assumption makes agents' behavior independent of a particular realization of the prior. Therefore, we summarize each learning algorithm via its Bayesian-expected rewards, not worrying about its detailed performance on particular realizations of the prior.  Such summarization is essential for formulating lucid and general analytic results, let alone proving them. It is unclear how to incorporate performance signals in a theoretically tractable model. For the numerical results, the \ExptsModel accounts for competition in a more direct way \gaedit{by positing a more realistic model of consumer behavior that does not require them to have direct information on the algorithms deployed or the bandit problem faced by the firm. It further allows us to separate reputation vs. data advantage and makes our model amenable to numerical simulations.} Indeed, simulating the intricate interplay of learning dynamics and Bayesian priors appears computationally intractable.

\asedit{Principals' utilities (and, accordingly, bandit rewards) are not discounted with time.%
\footnote{However, our main theoretical result on the greedy algorithm extends to time-discounted rewards, see Section~\ref{sec:theory-extensions}.}
 While much of the early work focuses on  time-discounted bandits \citep{Gittins-book11}, non-discounted  formulations are prevalent in the bandits literature over the past two decades \citep{slivkins-MABbook,LS19bandit-book}, and better correspond to practical deployments \citep[\eg][]{DS-arxiv}. Moreover, the distinctions between better and worse bandit algorithms are not as well-understood under time-discounting.}

\gaedit{Finally, the assumption that firms commit to the algorithms at the start of the game deserves further justification. We impose this commitment assumption since it is more realistic than considering dynamic strategies and algorithms are, by definition, a commitment to a set of rules. In many contexts where such exploration algorithms are typically deployed, consumers arrive at such a rapid volume that it it is infeasible to consider a firm dynamically adapting its algorithm to each consumer that arrives. Indeed, \cite{DS-arxiv} points out how difficult it is to deploy standard exploration algorithms in practice and such dynamic strategies are substantially more difficult to deploy in practice. While it is possible to consider share-adaptive algorithms that would still keep the commitment assumption but allow for indirect dynamic responses to market conditions, such algorithms do not exist currently in the literature but are a good avenue for future work.}

>>>>

%We investigate these questions with
\xhdr{Our model: competition game.} We consider a stylized duopoly model in which two firms (\emph{principals}) simultaneously engage in exploration and compete for users (\emph{agents}). Principals compete on quality rather than on prices. Agents arrive sequentially, and choose among the two principals according to some decision rule. The chosen principal chooses an action (\eg a list of web search results to show to the user), the user experiences this action, and reports a reward. Each principal only observes its own users. Principals commit to their strategies simultaneously and in advance, with objective to maximize the market share. We investigate several model variants within this framework, where we vary agents' decision rule and/or give a first-mover advantage to one of the principals. In all variants, agents have little or no information about other agents' choices and rewards.

%These two processes are interlinked, as exploration decisions are experienced by users and informed by their feedback. We need to specify several conceptual pieces: how the principals and agents interact, what is the machine learning problem faced by each principal, and what is the information structure. Each piece can get rather complicated in isolation, let alone jointly, so we strive for simplicity. Thus, the key features of our model are as follows:

%\begin{itemize}
%\textbf{(i)} A new agent arrives in each round $t=1,2, \ldots$, and chooses among the two principals. The principal chooses an action (\eg a list of web search results to show to the agent), the user experiences this action, and reports a reward. All agents have the same ``decision rule" for choosing among the principals given the available information.

%\textbf{(ii)} Each principal faces a basic and well-studied version of the multi-armed bandit problem: for each arriving agent, it chooses from a fixed set of actions  (a.k.a. \emph{arms}) and receives a reward drawn independently from a fixed distribution specific to this action. The reward distributions are initially unknown.
% (but can be estimated over time from the data).

%\textbf{(iii)} Principals simultaneously announce their bandit algorithms before round $1$, and cannot change them afterwards.  Each principal's objective is to maximize its market share (the fraction of users choosing this principal). Each principal only observes agents that chose this principal.
%, but may have access to each platforms' reputation score (more on this in    Section~\ref{sec:intro-discussion}).
%\end{itemize}

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a bandit algorithm; as such, it adjusts itself to the observed rewards, but not to the competition. This modeling choice reflects the reality of industrial applications, which build on a huge body of knowledge in machine learning. Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical. We draw on the machine learning literature to compare bandit algorithms to one another in the absence of competition, so that we can talk about \emph{better} bandit algorithms in a principled way.%
\footnote{Comparisons between bandit algorithms are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. Generally,  ``better" algorithms are better in the long run, but could be worse initially. More on this in Preliminaries.} One baseline is \emph{greedy algorithms} that do not explicitly explore; in the absence of competition, it is known to be terrible for a wide variety of problem instances.

@@

\xhdr{Theoretical results.}
%We endow agents with Bayesian rationality, a common modeling approach for a theoretical investigation.
We consider a basic Bayesian model. We posit that agents have a common Bayesian prior on reward distributions, and know the principals' algorithms. For simplicity, agents do not receive any information about the previous agents' choices and rewards. Each agent knows the round (s)he arrives in, computes the Bayesian-expected reward for each principal, and use these two numbers to decide which principal to choose. We refer to this variant as the \emph{\TheoryModel}. We focus on qualitative, asymptotic differences in the algorithms' performance.

@@

@@

@@

@@

@@

### 2020-09-03 journal paper/content/sec-model.tex
A pass on the intro: compressed the model & algorithms, expanded on dynamic vs static strategies. Moved "discussion of the model" to the model section.

>>>>

\subsection{Discussion}
\label{sec:discussion}

\ascomment{copied from the intro, to be expanded}

Our models are stylized in several important  respects. Firms compete only on the quality of service, rather than, say, pricing or the range of products. Agents are myopic: they do not worry about how their actions impact their future utility.\footnote{In particular, agents do not attempt to learn over time, to second-guess or game future agents, or to manipulate the principals' learning algorithms. This is arguably typical in practice, in part because one agent's influence tends to be relatively small.}  Various performance signals available to the users, from personal experience to word-of-mouth to consumer reports, are abstracted as persistent ``reputation scores", and further simplified to average performance over a time window. On the machine learning side, our setup is geared to distinguish between ``simple" vs. ``better" vs. ``smart" bandit algorithms; we are not interested in state-of-art algorithms for very realistic bandit settings.

We consider two extremes: a simple Bayesian model with full Bayesian rationality and no performance signals, and a simple frequentist model with reputation scores and no prior knowledge or beliefs. For the theoretical results, the ``no-performance-signals" assumption makes agents' behavior independent of a particular realization of the prior. Therefore, we summarize each learning algorithm via its Bayesian-expected rewards, not worrying about its detailed performance on particular realizations of the prior.  Such summarization is essential for formulating lucid and general analytic results, let alone proving them. It is unclear how to incorporate performance signals in a theoretically tractable model. For the numerical results, the \ExptsModel accounts for competition in a more direct way \gaedit{by positing a more realistic model of consumer behavior that does not require them to have direct information on the algorithms deployed or the bandit problem faced by the firm. It further allows us to separate reputation vs. data advantage and makes our model amenable to numerical simulations.} Indeed, simulating the intricate interplay of learning dynamics and Bayesian priors appears computationally intractable.

\asedit{Principals' utilities (and, accordingly, bandit rewards) are not discounted with time.%
\footnote{However, our main theoretical result on the greedy algorithm extends to time-discounted rewards, see Section~\ref{sec:theory-extensions}.}
 While much of the early work focuses on  time-discounted bandits \citep{Gittins-book11}, non-discounted  formulations are prevalent in the bandits literature over the past two decades \citep{slivkins-MABbook,LS19bandit-book}, and better correspond to practical deployments \citep[\eg][]{DS-arxiv}. Moreover, the distinctions between better and worse bandit algorithms are not as well-understood under time-discounting.}

\gaedit{Finally, the assumption that firms commit to the algorithms at the start of the game deserves further justification. We impose this commitment assumption since it is more realistic than considering dynamic strategies and algorithms are, by definition, a commitment to a set of rules. In many contexts where such exploration algorithms are typically deployed, consumers arrive at such a rapid volume that it it is infeasible to consider a firm dynamically adapting its algorithm to each consumer that arrives. Indeed, \cite{DS-arxiv} points out how difficult it is to deploy standard exploration algorithms in practice and such dynamic strategies are substantially more difficult to deploy in practice. While it is possible to consider share-adaptive algorithms that would still keep the commitment assumption but allow for indirect dynamic responses to market conditions, such algorithms do not exist currently in the literature but are a good avenue for future work.}
